Bandit Multiclass Linear Classification: Efficient Algorithms for the Separable Case

### 2019-01-21 ICML-submission/algorithm-for-strongly-linearly-separable-data.tex
Update algorithm-for-strongly-linearly-separable-data.tex
Corrected a typo
Update algorithm-for-strongly-linearly-separable-data.tex
Reformulated the mistake lower bound theorem, added some small fix to the remark, fixed some typos in the kernel algorithm theorem, changed frac to slash throughout, and shortened the "observe feedback" lines in the two algorithms

The idea behind
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
to use $K$ copies of the \textsc{Binary Perceptron} algorithm; see e.g.
\citet[Section 3.3.1]{Shalev-Shwartz-2012}. Each copy corresponds to one of the
classes. At each timestep $t$, upon seeing example $x_t$, copy $i$ makes a
binary prediction on whether $x_t$ belongs to class $i$ or not. If at least one
copy makes a positive prediction, the algorithm chooses its prediction as one of
the classes corresponding to a positive prediction. If all copies make negative
predictions, the algorithms makes a prediction uniformly at random from
$\cbr{1,\ldots,K}$.

\begin{algorithm}[h]
\caption{\textsc{Bandit Algorithm for Strongly Linearly Separable Examples}
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$, number of rounds $T$.}
\REQUIRE{Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Compute $S_t = \left\{ i ~:~ 1 \le i \le K, \ \ip{w_i^{(t)}}{x_t} \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\STATE{Observe feedback $z_t \in \{0,1\}$ \\ \qquad where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ \\ \qquad  for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + x_t$}
\label{line:pos-update}
\ENDIF
\ELSE
\STATE{Predict $\widehat y_t \in S_t$ chosen arbitrarily}
\STATE{Observe feedback $z_t \in \{0,1\}$ \\ \qquad where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ \\ \qquad for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\label{line:neg-update}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Mistake upper bound]
\label{theorem:strongly-separable-examples-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, $K$ be a positive
integer, $\gamma$ be a positive real number, $R$ be a non-negative real number.
If the examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in V \times
\{1,2,\dots,K\}$ are strongly linearly separable with margin $\gamma$ and for
all $t \in \cbr{1,\ldots,T}$ satisfy $\norm{x_t} \le R$ then the expected number
of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:strongly-separable-examples-mistake-lower-bound}
Let $\gamma$ be a positive real number, and $R$ be a non-negative real
number. For any (possibly randomized) algorithm $\calA$ for the \textsc{Online
Multiclass Classification with Bandit Feedback} problem with $K \le
(R/\gamma)^2$ classes, there exists an inner product space $(V,
\ip{\cdot}{\cdot})$, a non-negative integer $T$ and a sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ in $V \times
\{1,2,\dots,K\}$ that are strongly linearly separable with margin $\gamma$, the
norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ and the expected
number of mistakes $\calA$ makes is at least $\frac{K-1}{2} \left\lfloor
\frac{1}{4} (R/\gamma)^2 \right\rfloor$.
\end{theorem}

\paragraph{Remark.} Irrespective of any conditions on $K$, $R$, and $\gamma$, a
trivial lower bound on the expected number of mistakes of any randomized
algorithm is $\frac{K-1}{2}$. To see this, note that the adversary can choose an
example $(R e_1, y)$, where $y$ is a label chosen uniformly from
$\cbr{1,2,\dots,K}$, and show this example $K$ times. The sequence of examples
trivially satisfies the strong linear separability condition, and the
$\frac{K-1}{2}$ expected mistake lower bound follows from Claim 2
of~\citet{Daniely-Helbertal-2013}. However, it is unclear if the trivial lower
bound is the best possible if $K$ is large, i.e., $\omega((R/\gamma)^2)$. We
leave this as an open problem.

@@

\begin{theorem}[Mistake upper bound for kernelized algorithm]
\label{theorem:kernelized-upper-bound}
Let $X$ be a non-empty set, let $(V, \ip{\cdot}{\cdot})$ be an inner product
space. Let $\phi:X \to V$ be a feature map and let $k:X \times X \to \R$,
$k(x,x') = \ip{\phi(x)}{\phi(x')}$ the associated positive positive definite
kernel. Let $K$ be a positive integer, $\gamma$ be a positive real number, $R$
be a non-negative real number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in
X \times \{1,2,\dots,K\}$ are labeled examples such that:
\begin{enumerate}
\item the mapped examples $(\phi(x_1), y_1)$, $(\phi(x_2), y_2)$, $\dots$, $(\phi(x_T), y_T)$
are strongly linearly separable with margin $\gamma$,
\item for all $t$ in $\cbr{1,\ldots,T}$, $k(x_t,x_t) \le R^2$,
\end{enumerate}
then the expected number of mistakes Algorithm~\ref{algorithm:kernelized} makes
is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

\begin{algorithm}[h]
\caption{\textsc{Kernelized Bandit Algorithm}
\label{algorithm:kernelized}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$, number of rounds $T$.}
\REQUIRE{Kernel function  $k(\cdot, \cdot)$.}
\STATE{Initialize $J_1^{(1)} = J_2^{(1)} = \dots = J_K^{(1)} = \emptyset$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$.}
\STATE{Compute \\ $S_t = \left\{ i ~:~ 1 \le i \le K, \ \sum_{(x,y) \in J_i^{(t)}} y k(x, x_t) \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\STATE{Observe feedback $z_t \in \{0,1\}$ \\ \qquad where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $J_i^{(t+1)} = J_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $J_i^{(t+1)} = J_i^{(t)}$ \\ \qquad  for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $J_{\widehat y_t}^{(t+1)} = J_{\widehat y_t}^{(t)} \cup \cbr{(x_t, +1)}$}
\ENDIF
\ELSE
\STATE{Predict $\widehat y_t \in S_t$ chosen arbitrarily}
\STATE{Observe feedback $z_t \in \{0,1\}$ \\ \qquad where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $J_i^{(t+1)} = J_i^{(t)}$ \\ \qquad for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $J_{\widehat y_t}^{(t+1)} = J_{\widehat y_t}^{(t)} \cup \cbr{(x_t, -1)}$}
\ELSE
\STATE{Set $J_i^{(t+1)} = J_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

>>>>

The idea behind
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
to use $K$ copies of the \textsc{Binary Perceptron} algorithm; see e.g.
\citet[Section 3.3.1]{Shalev-Shwartz-2012}. Each copy corresponds to one of the
classes. At each timestep $t$, upon seeing example $x_t$, copy $i$ makes a
binary prediction on whether $x_t$ belongs to class $i$ or not. If at least one
copy makes a positive prediction, the algorithm chooses its prediction as one of
the classes corresponding to a positive prediction. If all copies make negative
predictions, the algorithm makes a prediction uniformly at random from
$\cbr{1,\ldots,K}$.

\begin{algorithm}[h]
\caption{\textsc{Bandit Algorithm for Strongly Linearly Separable Examples}
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$, number of rounds $T$.}
\REQUIRE{Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Compute $S_t = \left\{ i ~:~ 1 \le i \le K, \ \ip{w_i^{(t)}}{x_t} \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\STATE{Observe feedback $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ \\ \qquad  for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + x_t$}
\label{line:pos-update}
\ENDIF
\ELSE
\STATE{Predict $\widehat y_t \in S_t$ chosen arbitrarily}
\STATE{Observe feedback $z_t  = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ \\ \qquad for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\label{line:neg-update}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Mistake upper bound]
\label{theorem:strongly-separable-examples-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, $K$ be a positive
integer, $\gamma$ be a positive real number, $R$ be a non-negative real number.
If the examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in V \times
\{1,2,\dots,K\}$ are strongly linearly separable with margin $\gamma$ and 
%%%%%
$\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
%for all $t \in \cbr{1,\ldots,T}$ satisfy $\norm{x_t} \le R$ 
%%%%%
then the expected number
of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:strongly-separable-examples-mistake-lower-bound}
Let $\gamma$ be a positive real number, and $R$ be a non-negative real
number. 
%%%%%
Any (possibly randomized) algorithm makes at least 
$((K-1)/2)\left\lfloor (R/\gamma)^2/4 \right\rfloor$ mistakes in expectation 
on some \textsc{Online
Multiclass Classification with Bandit Feedback} problem with $K \le
(R/\gamma)^2$ classes and labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in V \times
\{1,2,\dots,K\}$ for some inner product space $(V,
\ip{\cdot}{\cdot})$ that are strongly linearly separable with margin $\gamma$ and satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ .
%
%For any (possibly randomized) algorithm $\calA$ for the \textsc{Online
%Multiclass Classification with Bandit Feedback} problem with $K \le
%(R/\gamma)^2$ classes, there exists an inner product space $(V,
%\ip{\cdot}{\cdot})$, a non-negative integer $T$ and a sequence of labeled
%examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ in $V \times
%\{1,2,\dots,K\}$ that are strongly linearly separable with margin $\gamma$, the
%norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ and the expected
%number of mistakes $\calA$ makes is at least $\frac{K-1}{2} \left\lfloor
%\frac{1}{4} (R/\gamma)^2 \right\rfloor$.
%%%%%
\end{theorem}

\paragraph{Remark.} 
%%%%%
If $\gamma \le R$ then, irrespective 
%%%%%
of any other conditions on $K$, $R$, and $\gamma$, 
a trivial lower bound on the expected number of mistakes of any randomized
algorithm is $(K-1)/2$. To see this, note that the adversary can choose an
example $(R e_1, y)$, where $e_1$ is some arbitrary unit vector in $V$ and $y$ is a label chosen uniformly from
$\cbr{1,2,\dots,K}$, and show this example $K$ times. The sequence of examples
trivially satisfies the strong linear separability condition, and the
$(K-1)/2$ expected mistake lower bound follows from ~\citep[][Claim 2]{Daniely-Helbertal-2013}. However, it is unclear if the trivial lower
bound is the best possible if $K$ is large, i.e., $\omega((R/\gamma)^2)$. We
leave this as an open problem.

@@

\begin{theorem}[Mistake upper bound for kernelized algorithm]
\label{theorem:kernelized-upper-bound}
Let $X$ be a non-empty set, let $(V, \ip{\cdot}{\cdot})$ be an inner product
space. Let $\phi:X \to V$ be a feature map and let $k:X \times X \to \R$,
$k(x,x') = \ip{\phi(x)}{\phi(x')}$ be the associated positive definite
kernel. Let $K$ be a positive integer, $\gamma$ be a positive real number, $R$
be a non-negative real number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in
X \times \{1,2,\dots,K\}$ are labeled examples such that:
\begin{enumerate}
\item the mapped examples $(\phi(x_1), y_1)$, $(\phi(x_2), y_2)$, $\dots$, $(\phi(x_T), y_T)$
are strongly linearly separable with margin $\gamma$,
\item for all $t$ in $\cbr{1,\ldots,T}$, $k(x_t,x_t) \le R^2$,
\end{enumerate}
then the expected number of mistakes Algorithm~\ref{algorithm:kernelized} makes
is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

\begin{algorithm}[h]
\caption{\textsc{Kernelized Bandit Algorithm}
\label{algorithm:kernelized}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$, number of rounds $T$.}
\REQUIRE{Kernel function  $k(\cdot, \cdot)$.}
\STATE{Initialize $J_1^{(1)} = J_2^{(1)} = \dots = J_K^{(1)} = \emptyset$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$.}
\STATE{Compute \\ $S_t = \left\{ i ~:~ 1 \le i \le K, \ \sum_{(x,y) \in J_i^{(t)}} y k(x, x_t) \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\STATE{Observe feedback $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $J_i^{(t+1)} = J_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $J_i^{(t+1)} = J_i^{(t)}$ \\ \qquad  for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $J_{\widehat y_t}^{(t+1)} = J_{\widehat y_t}^{(t)} \cup \cbr{(x_t, +1)}$}
\ENDIF
\ELSE
\STATE{Predict $\widehat y_t \in S_t$ chosen arbitrarily}
\STATE{Observe feedback $z_t =  \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $J_i^{(t+1)} = J_i^{(t)}$ \\ \qquad for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $J_{\widehat y_t}^{(t+1)} = J_{\widehat y_t}^{(t)} \cup \cbr{(x_t, -1)}$}
\ELSE
\STATE{Set $J_i^{(t+1)} = J_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}
